---
title: "Assignment 1: The cocktail bar (data transformation and manipulation)"
author: "Eszter Palócz"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r}
# Set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com"))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(dplyr)
library(knitr)
library(stringr)
```

# Background

Your Cuban uncle sets up a cocktail bar in downtown Budapest. He has a secret list of cocktails that he would like to serve in the bar. He asks you to do create a few lists and tables to set up the shop. As your uncle is a secret hipster, he has a dataset on Github that you can use for the task!
  
## 1. Dataset

The cocktail_bar dataset contains information about cocktails and their ingredients. 

```{r}
url <- "https://raw.githubusercontent.com/nthun/cocktail-balance/refs/heads/master/cocktail_data.tsv"

cocktail_bar <- read.delim(url, sep = "\t")

str(cocktail_bar)
```

## 2. Transforming the data table and clean the ingredient variable

The ingredients are currently in a single cell for each cocktail. 
In this section the ingredients will be placed in separate rows and will be 
cleaned of all quantities (e.g. 1/2 oz, 2 dashes, etc.), and the indicator of 
alcohol content (e.g. 47.3% abv).

```{r}
# Putting ingredients in separate rows
cocktailbar_new <- cocktail_bar %>% 
    separate_rows(ingredients, sep = "<br/>|<br>|<b4/>")

# Removing quantities, alcohol content, unnecessary space 
cocktailbar_new <- cocktailbar_new %>%
  mutate(
    ingredients = str_remove_all(ingredients,"\\b[0-9]+(\\/[0-9]+)?\\s*(oz|dash|dashes|drops|bsp)\\b"),
    ingredients = str_remove_all(ingredients, "\\b[0-9]+\\b"),
    ingredients = str_replace_all(ingredients, "\\.", ""),
    ingredients = str_replace_all(ingredients, "\\s+", ""),
    ingredients = str_remove_all(ingredients, "\\(.*%abv\\)")
  )
```

## 3. Arranging all ingredients in alphabetical order

Before opening the bar, it is important to find a reliable supplier that has all
the ingredients. The owner needs to send a list of all possible ingredients.
The supplier does not need the quantities just the names of the ingredients.

```{r}
# Arranging the ingredients in alphabetical order
cocktailbar_new <- cocktailbar_new %>%
  arrange(ingredients)
```

## 4. Number of unique ingredients

How many different ingredients you will the owner need?

```{r}
# Counting unique ingredients 
unique_ingredients <- cocktailbar_new %>%
  count(ingredients, sort = TRUE) 
  
print(unique_ingredients)
```

There are 62 unique ingredients needed for the cocktails.

## 5. What are the top 10 ingredients?

What are the 10 most frequently used ingredients? 

```{r}
# Counting the frequency of each ingredient
ingredient_freq <- cocktailbar_new %>%
  group_by(ingredients) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# top 10 most frequently used ingredients 
top_ingredients <- mutate(unique_ingredients, rank = dense_rank(desc(n))) %>% 
  filter(rank <= 10)                        

# top 10 ingredients (including ties): 
kable(top_ingredients)
```

## 6. Which cocktail(s) has/have the most ingredients?

Count the number of ingredients and filter all the cocktails that has that many.

```{r}
# Count the number of ingredients for each cocktail
number_of_ingredients <- cocktailbar_new %>%
  group_by(name) %>%
  summarise(num_ingredients = n_distinct(ingredients)) %>%
  ungroup()

# Maximum number of ingredients in one cocktail
max_ingredients <- max(number_of_ingredients$num_ingredients)

# All cocktails that have the maximum number of ingredients
cocktails_max_ingredients <- number_of_ingredients %>%
  filter(num_ingredients == max_ingredients)

# Finding the cocktails with the maximum number of ingredients
kable(cocktails_max_ingredients)
```

## 7. How many ingredients appear in only one cocktail (rare ingredient)?

```{r}
# Finding rare ingredients 
rare_ingredients <- cocktailbar_new %>%
  group_by(ingredients) %>%
  summarise(cocktail_count = n_distinct(name)) %>%
  filter(cocktail_count == 1)  # Keep only ingredients appearing in one cocktail

# Count the number of rare ingredients
number_rare_ingredients <- nrow(rare_ingredients)

# The number of rare ingredients
print(number_rare_ingredients)
```

The number of rare ingredients that appear in only one cocktail is 28. 


## 8. Which cocktail has an ingredient that is only used in one cocktail?

```{r}
# Step 1: Count the occurrence of each ingredient
ingredient_counts <- cocktailbar_new %>%
  count(ingredients) %>%  
  filter(n == 1)

# Step 2: Find the cocktails that contain those rare ingredients
cocktails_with_rare_ingredients <- cocktailbar_new %>%
  filter(ingredients %in% ingredient_counts$ingredients)

# Step 3: Select relevant columns (cocktail name and ingredient) and display
one_cocktail_ingr <- cocktails_with_rare_ingredients %>%
  dplyr::select(name, ingredients) %>%
  distinct(name)  

# Show the result
kable(one_cocktail_ingr)
```

## 9. What are the cocktails without rare ingredients?

```{r}
cocktails_without_rare_ingredients <- cocktailbar_new %>%
  filter(!ingredients %in% rare_ingredients$ingredients) %>%
  distinct(name)

kable(cocktails_without_rare_ingredients)
```

There are 55 cocktails without rare ingredients. 

## 10. A cheat sheet for the bartender

The following matrix shows all cocktail names as rows and all ingredients as 
columns. When a cocktail requires an ingredient, there is an "X" in the cell, 
otherwise, the cell is empty.

```{r}
# Creating a unique list of cocktails and ingredients
unique_cocktails <- unique(cocktailbar_new$name)
unique_ingredients <- unique(cocktailbar_new$ingredients)

# Creating an empty matrix
cocktail_matrix <- matrix("", nrow = length(unique_cocktails), ncol = length(unique_ingredients))

# Setting row and column names
rownames(cocktail_matrix) <- unique_cocktails
colnames(cocktail_matrix) <- unique_ingredients

# Filling the matrix based on the data
for (i in 1:nrow(cocktailbar_new)) {
  cocktail_index <- match(cocktailbar_new$name[i], unique_cocktails)
  ingredient_index <- match(cocktailbar_new$ingredients[i], unique_ingredients)
  cocktail_matrix[cocktail_index, ingredient_index] <- "X"
}

# Printing the matrix
kable(cocktail_matrix)
```
