---
title: 'Assignment 2: Data visualization'
author: "Eszter Palócz"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
# Set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com"))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
install.packages("tidytuesdayR")
library(tidytuesdayR)
```

## Task 1: Climbing expeditions
### Dowloading and installing the necessary packages and data 
```{r}
#Getting the data 
tuesdata <- tidytuesdayR::tt_load('2020-09-22')
expeditions <- tuesdata$expeditions
```

```{r}
# Installing necessary packages
install.packages("dplyr")
install.packages("ggplot2")
install.packages("tidyr")

# Loading the necessary libraries
library(tidytuesdayR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(forcats)
```

### Exploring the data 
```{r}
# Viewing the first few rows of the data
head(expeditions)

# Basic summary statistics
summary(expeditions)

# Checking for missing data
sum(is.na(expeditions))
```

### Manipulating the data for the plot
```{r}
# Apply fct_lump to retain only the 15 most frequent peaks
expeditions <- expeditions %>%
  mutate(peak_name = fct_lump(peak_name, n = 15))

# Check the levels of the 'peak' column
levels(expeditions$peak_name)

# Remove the "Other" category
expeditions <- expeditions %>%
  filter(peak_name != "Other")

# View the updated data
head(expeditions)
```

### Further manipulations
```{r}
# Count the number of expeditions for each peak
peak_counts <- expeditions %>%
  group_by(peak_name) %>%
  summarise(expedition_count = n(), .groups = "drop")

print(peak_counts)

# Reorder the peaks based on the sum of expeditions for each peak using fct_reorder()

expeditions <- expeditions %>%
  mutate(peak_name = fct_reorder(peak_name, peak_counts$expedition_count[match(peak_name, peak_counts$peak_name)]))

levels(expeditions$peak_name)

expeditions %>%
  arrange(desc(peak_name))
```

### Creating the plot 
```{r}
# Plot the data 
install.packages("viridis")
library(viridis)

expeditions %>%
  ggplot(aes(y = peak_name, fill = season)) +
  geom_bar() +
  scale_fill_viridis(discrete = TRUE) + 
  labs(title = "The 15 most popular peaks stacked by season of expedition", x = "Number of exhibitions", y = "") +
  theme_light() +
  theme(legend.position = "bottom") 
```


## Task 2: PhDs awarded
### Dowloading the data from Tidy Tuesday
```{r}
#Loading the data 
tuesdata <- tidytuesdayR::tt_load('2019-02-19')

phddata <- tuesdata$phd_by_field
```

### Checking the data 
```{r}
# Viewing the first few rows of the data and the dataset
head(phddata)

str(phddata)

# Basic summary statistics
summary(phddata)

# Checking for missing data
sum(is.na(phddata))
```

### Data manipulations for the plot
```{r}
# Aggragate the data to count all PhDs by broad field
str(phddata)

sum_broadfield <- phddata %>%
  mutate(n_phds = as.numeric(n_phds)) %>%  # Convert n_phds to numeric
  group_by(broad_field, year) %>%
  summarise(n_phds = sum(n_phds, na.rm = TRUE), .groups = "drop")
```

### Creating the plot
```{r}
# Plot
library(scales)
install.packages("RColorBrewer")
library(RColorBrewer)

ggplot(sum_broadfield, aes(x = year, y = n_phds, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma_format()) +
  labs(title = "Total Number of Ph.D.-s in the US by Year",
       x = "",
       y = "") +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")
```


## Task 3: Commute in the US
### Dowloading the data from Tidy Tuesday and checking the data structure
```{r}
#Loading the data 
tuesdata <- tidytuesdayR::tt_load('2019-11-05')

commute <- tuesdata$commute

# Checking the data structure 
str(commute)
```

### Data manipulations for the plot
```{r}
# Aggregate the data to count all commutes by state
commute_by_state <- commute %>%
  group_by(state) %>%
  summarise(total_commutes = n(), .groups = "drop")

# Checking the first rows of the data 
head(commute_by_state)

# Number of people walking and biking to work by state
library(scales)

commute_filtered <- commute %>%
  filter(mode %in% c("Walk", "Bike")) %>%
  group_by(state, mode, state_region, state_abb) %>%
  summarise(total_people = sum(n, na.rm = TRUE), .groups = "drop") %>%
  spread(key = mode, value = total_people, fill = 0)
```

### Creating the plot 
```{r}
# Creating the plot 

ggplot(commute_filtered, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  scale_x_log10(labels = comma_format()) +  
  scale_y_log10(labels = comma_format()) +  
  geom_text(aes(label = state_abb), hjust = 0.5, vjust = 0.5, size = 3, check_overlap = TRUE, color = "black") +
  labs(
    x = "Number of ppl walking to work (log N)", 
    y = "Number of people biking to work (log N)", 
    title = "Number of people walking vs. biking to work in each USA state") +
  theme_light()
```